Reducing Linearizability of Priority Queues and More Data Structures to State Reachability

### 2017-07-13 co-regular-EPQ-arxiv.tex
arxiv

We show that the set of executions for which some projection fails the test $\textit{Check-PQ-Conc-NonRec}$ can be characterized using a set of register automata, modulo a value renaming. The possibility of renaming values (which is complete for checking data independent implementations) allows to simplify the reasoning about projections. Thus, we assume that all the operations which are not in the projection failing this test use the same distinguished value $\top$, different from those used in the projection. Then, it is enough to find an automata characterization for the set of executions $e$ for which $\textit{Check-PQ-Conc-NonRec}$ fails, or equivalently, for which one
%a projection $e'$ fails $\textit{Check-PQ-Conc-NonRec}$ when one
of the following three formulas is false:
\begin{align*}
\hspace{-5mm}
\Gamma(e) := \mathsf{Has\text{-}}\Gamma(e) \Rightarrow \exists \alpha.\ \Gamma\mathsf{\text{-}Conc}(e,\alpha)\mbox{ with $\Gamma\in \{\mathsf{EmptyRemove}$, $\mathsf{UnmatchedMaxPriority}$, $\mathsf{MatchedMaxPriority}\}$}
\end{align*}
Intuitively, $\Gamma(e)$ states that $e$ is linearizable w.r.t. the set of sequential executions described by $\Gamma\mathsf{\text{-}Seq}$ (provided that $\mathsf{Has\text{-}}\Gamma(e)$ holds). Therefore, by an abuse of terminology, an execution $e$ satisfying $\Gamma(e)$ is called \emph{linearizable w.r.t. $\Gamma$}, or \emph{$\Gamma$-linearizable}.
Extending the automaton characterizing executions which are not $\Gamma$-linearizable, with self-loops that allow any operation with parameter $\top$ results in an automaton satisfying the following property called \emph{$\Gamma$-completeness}.

@@

$A \cap \mathcal{I} \neq \emptyset$ if and only if there exists $ e \in \mathcal{I}$ and $e' \in \textit{proj}(e)$ such that $e'$ is not $\Gamma$-linearizable.
\end{definition}

Section~\ref{ssec:aut} describes a $\mathsf{MatchedMaxPriority}$-complete automaton, the other automata being defined in Appendix~\ref{sec:appendix lemma and register automata for FIFO of single-priority executions} %Appendix \ref{sec:appendix proof and definition in section co-regular of EPQ1Lar},
%Appendix \ref{sec:appendix proof and definition in section co-regular of EPQ1Equal}
and Appendix \ref{subsec:co-regular of EPQ3}. Therefore, the following holds.

\begin{lemma}
There exists a $\Gamma$-complete automaton for each $\Gamma\in \{\mathsf{EmptyRemove}$, $\mathsf{UnmatchedMax}$ $\mathsf{Priority}$, $\mathsf{MatchedMaxPriority}\}$.
\end{lemma}

When defining $\Gamma$-complete automata, we assume that every implementation $\mathcal{I}$ behaves correctly, i.e., as a FIFO queue, when only values with the same priority are observed. More precisely, we assume that for every execution $e\in\mathcal{I}$ and every priority $p\in\mathbb{P}$, the projection of $e$ to values with priority $p$ is linearizable (w.r.t. $\seqPQ$). This property can be checked separately using register automata obtained from the finite automata in~\cite{DBLP:conf/icalp/BouajjaniEEH15} for FIFO queue (see Appendix~\ref{sec:appendix lemma and register automata for FIFO of single-priority executions} for more details). Note that this assumption excludes some obvious violations, such as a $\textit{rm}(a)$ operation happens before a $\textit{put}(a,p)$ operation, for some $p$.

Also, we consider $\Gamma$-complete automata for $\Gamma\in \{\mathsf{UnmatchedMaxPriority}, \mathsf{MatchedMaxPriority}\}$, recognizing executions which contain only one maximal priority. This is possible because any data-differentiated execution for which $\Gamma(e)$ is false has such a projection.
%Also, we consider $\Gamma$-complete automata for $\Gamma\in \{\mathsf{UnmatchedMaxPriority}, \mathsf{MatchedMaxPriority}\}$, recognizing executions where any two occurring priorities are comparable (w.r.t. $\prec$). This is possible because any data-differentiated execution which is not $\Gamma$-linearizable has such a projection.
Formally, given a data-differentiated execution $e$ and $p$ a maximal priority in $e$, $e\vert_{\preceq p}$ is the projection of $e$ to the set of values with priorities smaller than $p$. Then,

\begin{restatable}{lemma}{priExecutionIsEnough}
\label{lemma:pri execution is enough}
Let $\Gamma\in \{\mathsf{UnmatchedMaxPriority}, \mathsf{MatchedMaxPriority}\}$ and $e$ a data-differentiated execution. Then, $e$ is $\Gamma$-linearizable iff $e\vert_{\preceq p}$ is $\Gamma$-linearizable for some maximal priority $p$ in $e$.
\end{restatable}
\begin {proof} (Sketch)
%{\color {blue}We prove an equivalent property: $e$ is $\Gamma$-linearizable if and only if $e\vert_{\preceq p}$ is $\Gamma$-linearizable for some maximal priority $p$ in $e$.} %We consider the case of $\mathsf{MatchedMaxPriority}$ and the case of $\mathsf{UnmatchedMaxPriority}$ is similar.
To prove the $\textit{only if}$ direction, let $e$ be a data-differentiated execution linearizable w.r.t. $l = u \cdot \textit{put}(x,p) \cdot v \cdot \textit{rm}(x) \cdot w \in \mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}(s,x)$. Since $\mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}$ $(s,x)$ imposes no restriction on the operations in $u$, $v$ and $w$ with priorities incomparable to $p$, erasing all these operations results in a sequential execution which still satisfies this property. Similarly, for $\Gamma=\mathsf{UnmatchedMaxPriority}$.

The $\textit{if}$ direction follows from the fact that if the projection of an execution to a set of operations $O_1$ has a linearization $l_1$ and the projection of the same execution to the remaining set of operations has a linearization $l_2$, then the execution has a linearization which is defined as an interleaving of $l_1$ and $l_2$ (see Appendix~\ref{sec:appendix proof of Lemma pri execution is enough} for more details).

Thus, let $e$ be an execution such that $e\vert_{\preceq p}$ is linearizable w.r.t. $l = u \cdot \textit{put}(x,p) \cdot v \cdot \textit{rm}(x) \cdot w \in \mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}(s,x)$. By the property above, we know that $e$ has a linearization $l' = u' \cdot \textit{put}(x,p) \cdot v' \cdot \textit{rm}(x) \cdot w'$, such that the projection of $l'$ to values of priority comparable to $p$ is $l$.
%We can see that $u$, $v$ and $w$ are subsequences of $u'$, $v'$ an $w'$, respectively.
Since $\mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}(s,x)$ does not have a condition on values of priority incomparable to $p$, we obtain that $l' \in \mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}(s,\alpha)$.
\end {proof}

The following shows that $\Gamma$-complete automata enable an effective reduction of checking linearizability of concurrent priority queue implementations to state reachability. It is a direct consequence of the above definitions. Section~\ref{subsec:combine step-by-step linearizability and co-regular} discusses decidability results implied by this reduction.

\begin{restatable}{theorem}{ReduceEPQIntoStateReachability}
\label{lemma:reduce EPQ into state reachability}
Let $\mathcal{I}$ be a data-independent implementation, and $A(\Gamma)$ be a $\Gamma$-complete automaton for each $\Gamma$. Then,
$\mathcal{I} \sqsubseteq \seqPQ$ if and only if $\mathcal{I} \cap A(\Gamma) = \emptyset$ for all $\Gamma$.
\end{restatable}

%Section~\ref{ssec:aut} discusses

%When the data domain and priority domain are both fixed to be finite set, the register automata can be simulated with finite automata, and by Lemma \ref{lemma:reduce EPQ into state reachability}, checking linearizable w.r.t $\seqPQ$ is decidable.

%
%
% and
%
%Let $A(\Gamma)$ be a $\Gamma$-complete automaton for each $\Gamma$. The following theorem
%
%
%We will describe this automaton only in the case of $\Gamma=\mathsf{MatchedMaxPriority}$. The rest can be found in Appendix~\ref{}. These automata satisfy the following property
%
%the main technical challenge is to show that each of the following sets of executions can be characterized in terms of automata (modulo renaming of values):
%
%
% define a set of register automata whose product with a given implementation $\mathcal{I}$ accepts a linearizability violation whenever $\mathcal{I}$ is not linearizable.
%
%are precise enough to recognize every non-linearizable implementation.
%
%In this section, we divide checking linearizability w.r.t $\seqPQ$ into checking linearizablity w.r.t $\Gamma\mathsf{\text{-}Seq}(s,\alpha)$, each of which can be done in polynomial-time. Due to data-independence, violations to each $\Gamma\mathsf{\text{-}Seq}(s,\alpha)$ can be captured by register automata. This reduces checking linearizability w.r.t $\seqPQ$ into state reachability problem.
%
%
%
%\subsection{Co-Regular}
%\label{subsec:definition of co-regular}
%

%Let $\Gamma\mathsf{\text{-}Condition}(e)$ be $\mathsf{HasEmptyRemoves}(e)$, $\neg \mathsf{HasEmptyRemoves}(e) \wedge \mathsf{HasUnmatchedMaxPriority}(e)$, and $\mathsf{HasMatchedMaxPriority}(e)$, for $\Gamma = \mathsf{EmptyRemove}, \mathsf{UnmatchedMaxPriority}, \mathsf{MatchedMaxPriority}$, respectively. $\textit{Check-PQ-Conc}$ inspires us that, when $\Gamma\mathsf{\text{-}Condition}(e)$ holds, there should exist some $\alpha$, such that  $\Gamma\mathsf{\text{-}Conc}(e,\alpha)$ holds. The following lemma states this formally. Let $\textit{proj}(e)$ be the set of projections of $e$ into set of values. When refer to $\textit{proj}(e)$, we implicitly assume that each $\textit{rm}(\textit{empty})$ in $e$ has a ghost argument that is unique. We say that execution $l$ is linearizable w.r.t $\Gamma$, if we can find a sequence $s$ where $e \sqsubseteq s$ and $\Gamma\mathsf{\text{-}Seq}(s,\alpha)$ holds for some $\alpha$.

%\begin{restatable}{lemma}{EPQasMultiInMRpriforHistory}
%\label{lemma:EPQ as multi in MRpri for history}
%Given a data-differentiated execution $e$, $e \sqsubseteq \seqPQ$, if and only if, $\forall e' \in \textit{proj}(e)$, $\forall \Gamma\in \{\mathsf{EmptyRemove}$, $\mathsf{UnmatchedMaxPriority}$ $,\mathsf{MatchedMaxPriority}\}$,  $\Gamma\mathsf{\text{-}Condition}(e)$ $\Rightarrow$ $e$ is linearizable w.r.t $\Gamma$.
%\end{restatable}

%Note that due to data-independence, register automata are used to detect some renamed violations, instead of store and checking directly. Let us introduce the notion of co-regular, which reduce checking linearizable w.r.t $\Gamma\mathsf{\text{-}Conc}(e,\alpha)$ into checking set of witness automata.
%
%%\vspace{-6pt}
%\begin{definition}\label{def:co-regular of rules of extended priority queues}
%Given $\Gamma\in \{\mathsf{EmptyRemove}$, $\mathsf{UnmatchedMaxPriority}$ $,\mathsf{MatchedMaxPriority}\}$. $\Gamma$ is co-regular, if there is a finite set $\textit{Auts}_{\Gamma}$ of register automata, such that for each data-independent implementation $\mathcal{I}$, we have that:
%
%$\textit{Auts}_{\Gamma} \cap \mathcal{I} \neq \emptyset$ if and only if $\exists e \in \mathcal{I}_{\neq},e' \in \textit{proj}(e)$, such that $\Gamma\mathsf{\text{-}Condition}(e')$ holds but $e$ does not linearizable w.r.t $\Gamma$.
%
%$\seqPQ$ is co-regular, if each $\Gamma$ is co-regular.
%\end{definition}

@@

%{\color {blue}When $\mathsf{MatchedMaxPriority}$ is false for some execution $e$, then there are two possibilities:
%
%\begin{itemize}
%\setlength{\itemsep}{0.5pt}
%\item[-] There exists value $x$, let $e_x$ be obtained from $e$ when all the values of maximal priorities in $e$ but $x$ are removed. Then $\mathsf{MatchedMaxPriority\text{-}Conc}(e_x,x)$ is false.
%
%\item[-] There exists set $X$ of values, let $e'$ be obtained from $e$ when all the values of maximal priorities in $e$ but values in $X$ are removed. Then, although $\mathsf{MatchedMaxPriority\text{-}Conc}(e'_x,x)$ is true for each $x \in X$, there is no value $y$ to make $\mathsf{MatchedMaxPriority\text{-}Conc}(e',y)$ to be true.
%\end{itemize}
%
%It is surprised to see that these two possibilities can be captured by two sub-cases of $\mathsf{MatchedMaxPriority}$, depending on whether is only one value $x$ with maximal priority $p$. We denote by $\mathsf{MatchedMaxPriority}^{>}$ the strengthening of $\mathsf{MatchedMaxPriority}$ with this condition (i.e., all the values other than $x$ have a priority strictly smaller than $p$) and $\mathsf{MatchedMaxPriority}^{=}$ for the strengthening of the same formula with the negation of this condition (i.e., there exists another value $y$ with priority $p$).
%}

A differentiated execution $e$ is not $\mathsf{MatchedMaxPriority}$-linearizable when all the $\textit{put}$ operations in $e$ using the maximal priority $p$ are matched, and $e$ is not linearizable w.r.t. the set of sequential executions satisfying $\mathsf{MatchedMaxPriority\text{-}Seq}(e,x)$ for each value $x$ of priority $p$. We consider two cases depending on whether $e$ contains exactly one value with priority $p$ or at least two values. We denote by $\mathsf{MatchedMaxPriority}^{>}$ the strengthening of $\mathsf{MatchedMaxPriority}$ with the condition that all the values other than $x$ have a priority strictly smaller than $p$ (corresponding to the first case), and by $\mathsf{MatchedMaxPriority}^{=}$ the strengthening of the same formula with the negation of this condition (corresponding to the second case).
We use particular instances of register automata~\cite{DBLP:journals/tcs/KaminskiF94,DBLP:conf/icalp/Cerans94,DBLP:conf/stacs/SegoufinT11} whose states include only two registers, one for storing a priority guessed at the initial state, and one for storing the priority of the current action in the execution. The transitions can check equality or the order relation $\prec$ between the values stored in the two registers. Instead of formalizing the full class of register automata, we consider a simpler class which suffices our needs. More precisely, we consider a class of labeled transition systems whose states consist of a finite control part and a register $r$ interpreted to elements of $\mathbb{P}$. The transition labels can be one of the following:
\begin{itemize}
	\item $r=*$ for storing an arbitrary value to $r$,
	\item $\textit{call}(\textit{rm},a)$ and $\textit{ret}(\textit{rm},a)$ for reading call/return actions of a remove,
	\item $\textit{call}(\textit{put},d,g)$ where $g\in\{=r,\prec r,true\}$ is a guard, for reading a call action $\textit{call}(\textit{put},d,p)$ of a put and checking whether $p$ is either equal to or smaller than the value stored in $r$, or arbitrary,
	\item $\textit{ret}(\textit{put},d,true)$ for reading a return action $\textit{ret}(\textit{put},d,p)$ for any $p$.
\end{itemize}
The set of words accepted by such a transition system can be defined as usual.

%$\textit{call}(\textit{put},d,g)$ $\textit{ret}(\textit{put},d,g),$ $\textit{call}(\textit{rm},d),\textit{ret}(\textit{rm},d)$, where $d \in \mathbb{D} \cup \{ \textit{empty} \}$, and $g$ and $\top$ are guards
%
% which are complete for each case. Since we don't require the full generality of register automata,
%
%
%We define $\Gamma$-complete automata for both cases and their union will give a $\mathsf{MatchedMaxPriority}$-complete automaton
%
%
%
%Register automata is a kind of finite automata with registers used to deal with infinite data domain, where register can be assigned values and do simple operations. In this paper we use a restricted kind of register automata, which contains only one register, whose value is obtained once by guessing and never changed. Our register machine contains a register $r$. Its transition labels are $r=*,\textit{call}(\textit{put},d,g), \textit{ret}(\textit{put},d,g),$ $\textit{call}(\textit{rm},d),\textit{ret}(\textit{rm},d)$, where $d \in \mathbb{D} \cup \{ \textit{empty} \}$, and $g$ and $\top$ are guards. The first transition from initial state is a $r=*$ transition, and is used to assign an arbitrary value to $r$. A guard is a predicate chosen from $\{=r,< r,\top \}$, which represents equals the value of $r$, less than the value of $r$, and always holds, respectively. Given an execution $e = \alpha_1 \cdots \alpha_k$ of priority queue, register automata $\mathcal{A}$ accepts $e$, if there exists value $d_r\in \mathbb{D}$ and transitions $q_0 \xrightarrow{\beta_1} q_1 \ldots \xrightarrow{\beta_k} q_k$ of $\mathcal{A}$ from a initial state to a final state, such that for each $i$
%
%\begin{itemize}
%\setlength{\itemsep}{0.5pt}
%\item[-] If $\alpha_i = \textit{call}(\textit{put},a,p)$, then $\beta_i= \textit{call}(\textit{put},a,g)$ and $g$ holds with $p$ and $d_r$. Similarly for $\textit{ret}(\textit{put},a,p)$.
%
%\item[-] Else, $\alpha_i = \beta_i$.
%\end{itemize}

@@

%We use two point to simply our proof of co-regular. The first point is to use the results in \cite{Bouajjani:2015} to ensure FIFO (first in first out) property of each single-priority projections. A execution where only values of one priority is putted is called a single-priority execution. It is obvious that each single-priority projection of sequences in $\seqPQ$ satisfy FIFO property. Let $\textit{toQueue}(e)$ be an execution generated from $e$ by transforming $\textit{put}$ and $\textit{rm}$ into $\textit{enq}$ and $\textit{deq}$, respectively, and then discarding priorities. In Appendix \ref{sec:appendix proof and definition in section definition of co-regular}, according to the result of queue in \cite{Bouajjani:2015}, we construct a set $\textit{Auts}_{\textit{sinPri}}$ of witness automata, and shows that they are enough to ensure that for each data-differentiated executions, each of its single-priority projection without $\textit{rm}(\textit{empty})$ to have ``FIFO'' property, as shown by the following lemma.
%
%\begin{restatable}{lemma}{AutoForEPQwithSignlePri}
%\label{lemma:automata for extended priority queue with single priority}
%
%Given a data-independent implementations $\mathcal{I}$ of extended priority queue, $\mathcal{I} \cap \textit{Auts}_{\textit{sinPri}} \neq \emptyset$, if and only if there exists $e \in \mathcal{I}_{\neq}$, $e' \in \textit{proj}(e)$, such that $e'$ is single-priority  without $\textit{rm}(\textit{empty})$, and $\textit{toQueue}(e')$ does not linearizable to queue.
%\end{restatable}
%
%According to Lemma \ref{lemma:automata for extended priority queue with single priority}, from now on, it is safe to assume that, for each data-differentiated execution, any of its single-priority projection without $\textit{rm}(\textit{empty})$ has ``FIFO'' property. This effectively exclude some obvious violations, such as $\textit{rm}(a)$ happens before $\textit{put}(a,\_)$.

%Lemma \ref{lemma:pri execution is enough} makes our co-regular proof more clear. One consequence of it is that for given $e\vert_{\preceq p}$, $\Gamma\mathsf{\text{-}Condition}(e')$ holds only for one $\Gamma$.
%
%
%
%
%
%
%\subsection{Co-Regular of $\mathsf{MatchedMaxPriority}^{>}$}
%\label{subsec:co-regular of EPQ1Lar}

>>>>

We show that the set of executions for which $\textit{Check-PQ-Conc-NonRec}$ fails on some projection can be described using register automata, modulo a value renaming. Renaming values (which is complete under data independence) allows to simplify the reasoning about projections. W.l.o.g., we assume that all the operations which are not in the projection failing this test use the same distinguished value $\top$, different from those in the projection. Then, it is enough to find an automata characterization of the executions $e$ for which $\textit{Check-PQ-Conc-NonRec}(e)$ is $\mathsf{false}$, i.e., for which
$\Gamma(e) := \mathsf{Has\text{-}}\Gamma(e) \Rightarrow \exists \alpha.\ \Gamma\mathsf{\text{-}Conc}(e,\alpha)$ is false, for some $\Gamma\in \{\mathsf{EmptyRemove}$, $\mathsf{UnmatchedMaxPriority}$, $\mathsf{MatchedMaxPriority}\}$.
Intuitively, $\Gamma(e)$ states that $e$ is linearizable w.r.t. the set of sequential executions described by $\Gamma\mathsf{\text{-}Seq}$ (provided that $\mathsf{Has\text{-}}\Gamma(e)$ holds). Therefore, by an abuse of terminology, an execution $e$ satisfying $\Gamma(e)$ is called \emph{linearizable w.r.t. $\Gamma$}, or \emph{$\Gamma$-linearizable}.
Extending the automaton characterizing non $\Gamma$-linearizable executions with self-loops that allow any operation with argument $\top$ results in an automaton satisfying the following property called \emph{$\Gamma$-completeness}.

@@

$A \cap \mathcal{I} \neq \emptyset$ iff there exists $e \in \mathcal{I}$ and $e' \in \textit{proj}(e)$ such that $e'$ is not $\Gamma$-linearizable.
\end{definition}

We can show that for any $\Gamma\in \{\mathsf{EmptyRemove}$,$\mathsf{UnmatchedMaxPriority}$,$\mathsf{MatchedMaxPriority}\}$ there exists a $\Gamma$-complete automaton. For lack of space, we only consider the case $\Gamma=\mathsf{MatchedMaxPriority}$ in
Section~\ref{ssec:aut}.
When defining $\Gamma$-complete automata, we assume that every implementation $\mathcal{I}$ behaves correctly, i.e., as a FIFO queue, when only values with the same priority are observed. More precisely, we assume that for every execution $e\in\mathcal{I}$ and every priority $p\in\mathbb{P}$, the projection of $e$ to values with priority $p$ is linearizable (w.r.t. $\seqPQ$). This property can be checked separately using register automata similar to the automata in~\cite{DBLP:conf/icalp/BouajjaniEEH15} describing FIFO queue violations (see Appendix~\ref{sec:appendix lemma and register automata for FIFO of single-priority executions} for more details). This assumption excludes some obvious violations, such as an $\textit{rm}(a)$ operation happening before a $\textit{put}(a,p)$ operation, for some $p$.

For $\Gamma\in \{\mathsf{UnmatchedMaxPriority}, \mathsf{MatchedMaxPriority}\}$, we consider $\Gamma$-complete automata recognizing executions which contain only one maximal priority. This is w.l.o.g. because any data-differentiated execution for which $\Gamma(e)$ is false has such a projection.
Formally, given a data-differentiated execution $e$ and $p$ a maximal priority in $e$, $e\vert_{\preceq p}$ is the projection of $e$ to the set of values with priorities smaller or equal to $p$. Then,

\begin{lemma}
\label{lemma:pri execution is enough}
For $\Gamma\in \{\mathsf{UnmatchedMaxPriority}, \mathsf{MatchedMaxPriority}\}$, a data-differentiated execution $e$
 is $\Gamma$-linearizable iff $e\vert_{\preceq p}$ is $\Gamma$-linearizable for some maximal priority $p$ in $e$.
\end{lemma}
\begin {proof} (Sketch)
For the ``only-if'' direction, let $e$ be a data-differentiated execution linearizable w.r.t. $l = u \cdot \textit{put}(x,p) \cdot v \cdot \textit{rm}(x) \cdot w$ s.t. $\mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}(l,x)$ holds. Since the predicate $\mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}(l,x)$ imposes no restriction on the operations in $u$, $v$, and $w$ with priorities incomparable to $p$, erasing all these operations results in a sequential execution which still satisfies this predicate. Similarly, for $\Gamma=\mathsf{UnmatchedMaxPriority}$.

The ``if'' direction follows from the fact that if the projection of an execution to a set of operations $O_1$ has a linearization $l_1$ and the projection of the same execution to the remaining set of operations has a linearization $l_2$, then the execution has a linearization which is defined as an interleaving of $l_1$ and $l_2$ (see Appendix~\ref{sec:appendix proof of Lemma pri execution is enough} for more details).

Thus, let $e$ be an execution such that $e\vert_{\preceq p}$ is linearizable w.r.t. $l = u \cdot \textit{put}(x,p) \cdot v \cdot \textit{rm}(x) \cdot w$ where $\mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}(l,x)$ holds. By the property above, we know that $e$ has a linearization $l' = u' \cdot \textit{put}(x,p) \cdot v' \cdot \textit{rm}(x) \cdot w'$, such that the projection of $l'$ to values of priority comparable to $p$ is $l$.
Since $\mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}(l,x)$ doesn't constrain the values of priority incomparable to $p$, we obtain that $\mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}(l',\alpha)$ also holds.
\end {proof}

We shows in Appendix \ref{subsec:appendix co-regular of EPQ2Lar} and Appendix \ref{subsec:appendix co-regular of EPQ2Equal} that it is safe to ignore constructing $\mathsf{UnmatchedMaxPriority}$-complete register automata, and we give the $\mathsf{EmptyRemove}$-complete register automata in Appendix \ref{subsec:co-regular of EPQ3}. The existence of $\Gamma$-complete automata enable an effective reduction of checking linearizability of concurrent priority queue implementations to state reachability.
Section~\ref{subsec:combine step-by-step linearizability and co-regular} discusses decidability results implied by this reduction.

\begin{theorem}
\label{lemma:reduce EPQ into state reachability}
Let $\mathcal{I}$ be a data-independent implementation. Then, there is a $\Gamma$-complete automaton $A(\Gamma)$ for each $\Gamma$. Moreover,
$\mathcal{I} \sqsubseteq \seqPQ$ iff $\mathcal{I} \cap A(\Gamma) = \emptyset$ for all $\Gamma$.
\end{theorem}

@@

A differentiated execution $e$ is not $\mathsf{MatchedMaxPriority}$-linearizable when all the $\textit{put}$ operations in $e$ using the maximal priority $p$ are matched, and $e$ is not linearizable w.r.t. the set of sequential executions satisfying $\mathsf{MatchedMaxPriority\text{-}Seq}(e,x)$ for each value $x$ of priority $p$. We consider two cases depending on whether $e$ contains exactly one value with priority $p$ or at least two values. We denote by $\mathsf{MatchedMaxPriority}^{>}$ the strengthening of $\mathsf{MatchedMaxPriority}$ with the condition that all the values other than $x$ have a priority strictly smaller than $p$ (corresponding to the first case), and by $\mathsf{MatchedMaxPriority}^{=}$ the strengthening of the same formula with the negation of this condition (corresponding to the second case).
We use particular instances of register automata~\cite{DBLP:journals/tcs/KaminskiF94,DBLP:conf/icalp/Cerans94,DBLP:conf/stacs/SegoufinT11} whose states include only two registers, one for storing a priority guessed at the initial state, and one for storing the priority of the current action in the execution. The transitions can check equality or the order relation $\prec$ between the values stored in the two registers. Instead of formalizing the full class of register automata, we consider a simpler class which suffices our needs. Thus, we consider a class of labeled transition systems whose states consist of a finite control part and a register $r$ interpreted to elements of $\mathbb{P}$. The transition labels are:
\begin{itemize}
	\item $r=*$ for storing an arbitrary value to $r$,
	\item $\textit{call}(\textit{rm},a)$ and $\textit{ret}(\textit{rm},a)$ for reading call/return actions of a remove,
	\item $\textit{call}(\textit{put},d,g)$ where $g\in\{=r,\prec r,true\}$ is a guard, for reading a call action $\textit{call}(\textit{put},d,p)$ and checking if $p$ is either equal to or smaller than the value stored in $r$, or arbitrary,
	\item $\textit{ret}(\textit{put},d,true)$ for reading a return action $\textit{ret}(\textit{put},d,p)$ for any $p$.
\end{itemize}
The set of sequences (executions) accepted by such a transition system is defined as usual.

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=0.4 \textwidth]{figures/PIC-HIS-INTRO-GAP-EPQ1L.pdf}
  \caption{An execution that is not $\mathsf{MatchedMaxPriority}^{>}$-linearizable. We represent each operation as a time interval whose left, resp., right, bound corresponds to the call, resp., return action.}
  \label{fig:introduce gap for EPQ1Lar}
\end{figure}

Figure~\ref{fig:introduce gap for EPQ1Lar} contains a typical example of an execution $e$ which is not $\mathsf{MatchedMaxPriority}^>$-linearizable,
where $p_1 \prec p_4$, $p_2 \prec p_4$, and $p_3 \prec p_4$.
Intuitively, this is a violation because during the whole execution of $\textit{rm}(b)$, the priority queue stores a smaller priority value (which should be removed before $b$). To be more precise, we define \emph{the interval of a value $x$} as the time interval from the return of a put $\textit{ret}(\textit{put},x,p)$ to the call of the matching remove $\textit{call}(rm,x)$, or to the end of the execution if such a call action doesn't exist. This represents the time interval in which a value is guaranteed to be stored into the priority queue. Concretely, for a standard indexing of actions in an execution, a time interval is a closed interval between the indexes of two actions in the execution.
In \figurename~\ref{fig:introduce gap for EPQ1Lar}, the interval of each value of priority smaller than $p_4$ is pictured as a dashed line. There is no sequence $l$ s.t. $e \sqsubseteq l$ and $\mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}(l,b)$ hold, since each time point from $\textit{call}(\textit{rm},b)$ to $\textit{ret}(\textit{rm},b)$ is included in the interval of a smaller priority value,
and $\textit{rm}(b)$ can't take effect in the interval of a smaller priority value.
To formalize this scenario we use the notion of \emph{left-right constraint} defined below.

\begin{definition}\label{def:left-right constraint for matched put and rm operations}
Let $e$ be a data-differentiated execution which contains only one maximal priority $p$, and only one value $x$ of priority $p$ (and no $\textit{rm}(\textit{empty})$ operations).
The \emph{left-right constraint of $x$} is the graph $G$ where:
\begin{itemize}
\item the nodes are the values occurring in $e$,
\item there is an edge from $d_1$ to $x$, if $\textit{put}(d_1,\_) <_{\textit{hb}} \textit{put}(x,p)$ or $\textit{put}(d_1,\_) <_{\textit{hb}} \textit{rm}(x)$,
\item there is an edge from $x$ to $d_1$, if $\textit{rm}(x)<_{\textit{hb}}\textit{rm}(d_1)$ or $\textit{rm}(d_1)$ does not exists,
\item there is an edge from $d_1$ to $d_2$, if $\textit{put}(d_1,\_) <_{\textit{hb}} \textit{rm}(d_2,\_)$.
\end{itemize}
\end{definition}

The execution in \figurename~\ref{fig:introduce gap for EPQ1Lar} is not $\mathsf{MatchedMaxPriority}^>$-linearizable because the left-right constraint of the maximal priority value $b$ contains the cycle $f \rightarrow d \rightarrow c \rightarrow b \rightarrow f$. The following lemma states that the presence of such a cycle is equivalent to \emph{non} $\mathsf{MatchedMaxPriority}^>$-linearizability, and it is proved in Appendix \ref{sec:appendix proof and definition in section co-regular of EPQ1Lar}:

\begin{lemma}
\label{lemma:Lin Equals Constraint for EPQ1Lar}
Let $e$ be a data-differentiated execution such that
$\mathsf{Has\text{-}MatchedMaxPriority}(e)$ holds, $p$ is the maximal priority in $e$, and $\textit{put}(x,p)$ and $\textit{rm}(x)$ are only operations with arguments of priority $p$ in $e$.
Then, $e$ is $\mathsf{MatchedMaxPriority}$-linearizable iff the left-right constraint of $x$ contains no cycle going through $x$.
\end{lemma}

When the left-right constraint contains a cycle $d_1 \rightarrow \ldots \rightarrow d_m \rightarrow x \rightarrow d_1$, for some $d_1$,$\ldots$,$d_n\in \mathbb{D}$, we say that $x$ is \emph{covered} by $d_1,\ldots,d_m$. The shape of an execution witnessing such a cycle (i.e., the alternation between call/return actions) can be identified using our class of automata, the only complication being the unbounded number of values $d_1$,$\ldots$,$d_n$. However, by data independence, whenever an implementation contains such an execution it also contains an execution where all the values $d_1$,$\ldots$,$d_n$ are renamed to the same value $a$, and $x$ is renamed to $b$. Therefore, our automata can be defined over a fixed set of values $a$, $b$, and $\top$ (recall that $\top$ is used for operations outside of the non-linearizable projection).

To define a $\mathsf{MatchedMaxPriority}^>$-complete automaton, we need to consider all the possible orders between the call/return actions of the $\textit{put}$/$\textit{rm}$ operations that add and respectively, remove the value $b$. The case where the put happens-before the remove (as in \figurename~\ref{fig:introduce gap for EPQ1Lar}) is pictured in \figurename~\ref{fig:automata APQ1Lar-1 in paper}. This automaton captures the three possible ways of ordering the first action $\textit{ret}(\textit{put},a,\_)$ w.r.t. the actions with value $b$, which are pictured in \figurename~\ref{fig:executions APQ1Lar-1 in paper}(a) (this action cannot occur after $\textit{call}(\textit{rm},b,\_)$ since $b$ must be covered by the $a$-s). The paths corresponding to these three possible orders are: $q_1 \rightarrow q_2 \rightarrow q_3 \ldots \rightarrow q_7$, $q_1 \rightarrow q_2 \rightarrow q_3 \ldots \rightarrow q_{10}$, and $q_1 \rightarrow q_9 \rightarrow q_{10} \ldots \rightarrow q_7$. \figurename~\ref{fig:executions APQ1Lar-1 in paper} lists the four possible orderings of the call/return actions of adding and removing $b$, and also possible orders of the first $\textit{ret}(\textit{put},a,\_)$ w.r.t the actions with value $b$. Each such ordering corresponds to an automaton similar to the one in \figurename~\ref{fig:automata APQ1Lar-1 in paper}, their union defining a $\mathsf{MatchedMaxPriority}^>$-complete automaton. In Appendix \ref{sec:appendix proof and definition in section co-regular of EPQ1Lar}, three register automata is constructed according to the cases of \figurename~\ref{fig:executions APQ1Lar-1 in paper} (b), (c) and (d), respectively.

\begin{figure}[t]
  \centering
  \includegraphics[width=.8\textwidth]{figures/PIC_HIS_PQ1Lar-fouCase.pdf}
  \caption{Orderings to be considered when defining a $\mathsf{MatchedMaxPriority}^>$-complete automaton.}
  \label{fig:executions APQ1Lar-1 in paper}
\end{figure}
